Kaluza--Klein Cosmology from five-dimensional Lovelock--Cartan Theory

### 2016-09-21 KKosmology.tex
drop a duplicated footnote
Changes by Criz and SdP

%% \author{Francisca \surname{Ram\'irez}.}
%% \affiliation{\UTFSM.}

%% --------- Abstract
\begin{abstract}
  We study the Kaluza--Klein dimensional reduction of the Lovelock--Cartan theory on a five-dimensional manifold, where the compact dimension has $S^1$ topology. We find cosmological solutions of the Friedmann--Robertson--Walker class for the effective theory. The torsional degrees of freedom induce a nonvanishing energy-momentum tensor in four dimensions. We find solutions describing expanding, contracting and bouncing universes. The model shows a dynamical compactification of the extra dimension in some regions of the parameter space. 
\end{abstract}

@@

In four dimensions, the Einstein--Hilbert action with cosmological constant is the most general theory which leads to second order field equations for the metric. In higher dimensions, particular combinations of higher order terms in the curvature can be added to the gravitational action, giving rise to second order field equations for the metric. The most general theory in arbitrary dimensions, which preserves the features of the four-dimensional Einstein--Hilbert, is described by the Lanczos--Lovelock action~\cite{Lanczos:1938sf,*Lovelock:1971yv}. Such a theory, has no ghosts~\cite{Zumino:1985dp} and it has the same degrees of freedom as the Einstein--Hilbert Lagrangian in arbitrary dimensions~\cite{Henneaux:1990au}.

@@

The Gauss--Bonnet term, in four dimensions, adds no dynamics to the metric, since it represents a topological invariant proportional to the Euler characteristic class, which can be written locally as a boundary term. In dimensions higher than four it contributes to the field equations. This Lagrangian was also identified as the low-energy correction for a spin two field in string theory~\cite{Zwiebach:1985uq}.

\hl{Los siguientes dos parrafos no cuadran con la tonica anterior, modificar?}
On the one hand, classically, wormholes must be supported by a kind energy compatible with the cosmological hypothesis. Their existence in extended models, implies the presence of degrees of freedom that could provide an explanation for the exotic matter/energy abundance in the Universe. In the Riemannian five-dimensional Lanczos--Lovelock theory, exact wormholes solutions have been found in vacuum~\cite{Dotti:2006cp,*Dotti:2007az}. The same kind of solutions were found with matter~\cite{Mehdizadeh:2015jra}, in higher order Lanczos--Lovelock models~\cite{Mehdizadeh:2015dta}, and in higher order compactified Lanczos--Lovelock theories with torsion~\cite{Canfora:2008ka}. Additionally, the compactification of higher Lanczos--Lovelock terms has been considered in Refs.~\cite{MuellerHoissen:1985mm,*MuellerHoissen:1989yv} and their cosmology in Refs.~\cite{MuellerHoissen:1985ij,Deruelle:1986iv,Deruelle:1989fj}.

On the other hand, the renaissance of the Riemann--Cartan geometries came through the pioneer works of Sciama~\cite{Sciama:1962} and Kibble~\cite{Kibble:1961ba} on Poincar\'e gauge theories (see also Refs.~\cite{Blagojevic:2002du,*Hehl:1976kj}). In such a framework, the torsion appears as the field strength of translations, sourced by the spin-current.
%In such a framework, the vielbein is interpreted as the gauge connection of translations and the torsion as its field strength. The formulation of gravity as a gauge theory has major benefits in the problem of a unified picture of interactions, since it allows to interpret it as arising through a localization procedure, where the spacetime is a Riemann--Cartan manifold. 
%Due to this, several authors have relaxed the torsion-free condition since, as E. Cartan argued, there is no a fundamental reason to assume a torsion-free condition {\it a priori}. 
Moreover, the vacuum predictions of its simplest formulation---the Einstein--Cartan theory---holds the experimental tests of GR. The cosmological consequences of non-Riemannian geometries has been widely studied in the literature (for a review, see Ref.~\cite{Puetzfeld:2004yg}). %For instance, the pioneer works of Kopczy\'nski showed how the introduction of a spinning dust, acting as a source of the curvature and torsion, avoids the appearance of cosmological singularities~\cite{Kopczynski:1972,Kopczynski:1973}. In the context of the early evolution of the Universe, the spin density can drive the inflationary phase within the Einstein--Cartan theory, along with the avoidance of singularities~\cite{Gasperini:1986mv}. It has been shown in Ref.~\cite{Shie:2008ms} that torsional degrees of freedom might produce the current phase of accelerated expansion of the Universe. Additionally, the nonminimal coupling of the Euler characteristic form with a scalar field in Riemann--Cartan spacetimes, allows the torsional degrees of freedom to behave as an exotic induced-matter content, customary for the phase of accelerated expansion~\cite{Toloza:2013wi}.

In the framework of the Kaluza--Klein theories in higher dimensional Riemann--Cartan geometries, phenomenology of the extra dimensional torsion was found in Ref.~\cite{Kalinowski:1980da}, metric dependent torsional models of extra dimensions was studied in~\cite{Shankar:2012vd}, along with its consequences in cosmology~\cite{Chen:2009ep}. Compatification of higher dimensional Brans--Dicke models with torsion was considered in Ref.~\cite{German:1993bq}. Torsion-free black-hole solutions were found in~\cite{Aros:2007nn}, for first order compactified gravity.

The extension of the Lanczos--Lovelock theories with nonvanishing torsion is known as Lovelock--Cartan~\cite{Mardones:1990qc}. In that framework, we present a new class of cosmological solutions in five dimensions, where the compact dimension is $S^1$. The theory admits a nonvanishing torsion in vacuum due to the presence of the Gauss--Bonnet term, in contrast to the five-dimensional Einstein--Cartan theory. \hl{This are our results, in the intro?}After the dimensional reduction, the extra degrees of freedom generate an energy density and pressure as four-dimensional matter fields. In some cases, the solutions avoid the appearance of initial singularities and drive the accelerated expansion of the Universe, while the radius of the compact manifold goes to zero.

This work is organized as follows: In Sec.~\ref{KK} we introduce the torsional degrees of freedom compatible with the Kaluza--Klein geometry, and fix our notation and conventions. In Sec.~\ref{5EGB}, we study the dynamics of the general Lovelock--Cartan action in five-dimensional spacetime and its dimensional reduction. In Sec.~\ref{cosmos}, we look for cosmological solutions of the Friedmann--Robertson--Walker class with nontrivial torsion. Conclusions and remarks are given in Sec.~\ref{conclusions}. We also incorporate a number of appendices to make this work a self-contained article.

\section{Kaluza--Klein Ansatz in Riemann--Cartan Spacetimes\label{KK}}

In the following we will consider $M_N$ to be a $N$-dimensional differential manifold. Every quantity defined on $M_N$ will be denoted by hats $\hat{x}$. Capital Greek characters (coordinate indices) and capital Latin characters (Lorentz indices) run over the $N$ dimensions, while lower case run in the $(N-1)$-dimensional reduced manifold.

The differential structure on $M_N$ is determined by two fields. The spin connection 1-form, $\hat{\omega}^{AB} = \hat{\omega}^{AB}{}_{\Gamma}\,\text{d}\hat{x}^\Gamma$, describes the affine structure of the $N$-dimensional manifold, and the vielbein 1-form, $\hat{e}^A=\hat{e}^{A}{}_{\Gamma}\,\text{d}\hat{x}^\Gamma$, defines the metric structure of the same manifold through the relation $\hat{g}_{\Gamma\Delta} = \hat{e}^{A}{}_{\Gamma}\hat{e}^{B}{}_{\Delta}\hat{\eta}_{AB}$. The vielbein is the mapping that relates coordinate indices with the Lorentz indices.
In our convention, both Lorentzian metrics \(\hat{\eta}\) and \(\eta\) have positive signature, and the Levi-Civita symbols are such that
\[\hat{\epsilon}_{a_1 \cdots a_{N-1} N} \equiv \epsilon_{a_1 \cdots a_{N-1}},\]
%% \[\hat{\epsilon}_{0 1 \cdots N-1} = {\epsilon}_{0 1 \cdots N-1} = 1,\]
which allows us to define the reduced symbol by fixing the last index. Additionally, all Riemannian fields (torsion-free) will be explicited by a tilde, as it was done in Eq.~\eqref{GB} for the curvature.
%% In our convention, the flat Lorentzian metric will be $\hat{\eta}_{AB}=\diag(-,+,\cdots,+)$, and the $N$-dimensional Levi-Civita symbol $\hat{\epsilon}_{A_1 \cdots A_N}$ is reduced by defining
%% \begin{align*}
%% \hat{\epsilon}_{a_1 \cdots a_{N-1} N} \equiv \epsilon_{a_1 \cdots a_{N-1}},
%% \end{align*} 
%% since the coordinates of $M_N$ run from $0$ to $N-1$. Our convention fixes $\hat{\epsilon}_{01...N}=1$. Hereon, Riemannian fields (torsion-free) will be explicited by a tilde, as it was done in Eqs.~\eqref{GB} for the curvature.

Due to the topology of the manifold, we can expand the depenence of the fields on the extra coordinate, $z$, in Fourier series as  
\begin{equation}
  \label{Fourier}
  \hat{\varrho}(\hat{x}^\Gamma)=\sum_n\varrho_{(n)}(x)e^{i n z},
\end{equation}
where $x$ denotes the coordinates of the $(N-1)$-manifold, collectively. Henceforth, we will focus in the $n=0$ mode of the expasion, also refered to as the low-energy sector.

@@

The Kaluza--Klein (KK) ansatz for the metric relies on the premise that the compact dimension of $M_N$ is orthogonal to the rest of the manifold at each point (see for example Ref.~\cite{Blagojevic:2002du,*Hehl:1976kj}). This leads to the following metric structure
\begin{equation}
  \hat{g}_{\Gamma\Delta} =
  \begin{pmatrix}
    g_{\gamma\delta} +\frac{\hat{g}_{\gamma z}\hat{g}_{\delta z}}{\hat{g}_{zz}}&\hat{g}_{\gamma z}\\
    \hat{g}_{z\delta} & \hat{g}_{zz}
  \end{pmatrix}
  =
  \begin{pmatrix}
    g_{\gamma\delta} + \phi A_\gamma A_\delta&\phi A_\gamma\\
    \phi A_{\delta} & \phi
  \end{pmatrix},
\end{equation}
and its inverse
\begin{equation}
  \hat{g}^{\Gamma\Delta}=
  \begin{pmatrix}
    g^{\gamma\delta}&-A^\gamma\\
    -A^{\delta} & \phi^{-1}+A^2
  \end{pmatrix}.
\end{equation}
It introduces a scalar field $\phi$ and a vector field $A_\mu$ as new gravitational degrees of freedom. The vielbein which holds this structure for the metric has the following form
\begin{equation}
  \label{Dvielbein}
  \hat{e}^A{}_{\Gamma} =
  \begin{pmatrix}
    \hat{e}^a{}_{\gamma}& 0\\
    \hat{e}^N{}_{\gamma} & \hat{e}^N{}_{z}
  \end{pmatrix}
  =
  \begin{pmatrix}
    e^a{}_{\gamma}& 0\\
    \sqrt{\phi}A_\gamma & \sqrt{\phi}
  \end{pmatrix},
\end{equation}
and its inverse, $\hat{E}_A{}^{\Gamma}$, defined such that $\hat{E}_A{}^{\Gamma}\hat{e}^A{}_{\Delta}=\delta^\Gamma_{\Delta}$, reads
\begin{equation}
  \label{Dinversevielbein}
  \hat{E}_A{}^{\Gamma} =
  \begin{pmatrix}
    \hat{E}_a{}^{\gamma}& 0\\
    \hat{E}_a{}^{z} & \hat{E}_N{}^{z}
  \end{pmatrix}
  =
  \begin{pmatrix}
    E_a{}^{\gamma}& 0\\
    -A_a & \sqrt{\phi}^{-1}
  \end{pmatrix},
\end{equation}
and it shows that the compact manifold $S^1$ has its own tangent space, $T_pS^1$, at each point $p\in M_N$, and is independent of $T_pM_{N-1}$ in the sense that they do not mix. %It seems reasonable to think that if we want two manifolds to be orthogonal to one another, they will not mix vectors after the tangent mapping.

@@

A Riemannian connection compatible with the \mbox{$N$-dimensional} vielbein~\eqref{Dvielbein}, is built under the premise that $\mbox{d}\hat{e}^A + \hat{\tilde{\omega}}^{A}{}_B \wedge \hat{e}^B = 0$. Thus, we find
\begin{equation}
  \label{DRiemannconnection}
  \begin{split}
    \hat{\tilde{\omega}}^{ab}&=\tilde{\omega}^{ab}-\frac{1}{2}\sqrt{\phi}F^{ab}\hat{e}^N,\\
    \hat{\tilde{\omega}}^{Na}&=\frac{1}{2}\sqrt{\phi}F^a_{\ \ l}e^l+\frac{1}{2}\partial^a\ln\phi\hat{e}^N,
  \end{split}
\end{equation}
where $\tilde{\omega}^{ab}$ is the Riemannian spin connection of the reduced manifold and $F_{ab}$ are the components of the field strength of $A_b$ defined as
\begin{equation}
  F=\text{d}A=\frac{1}{2}F_{ab}\, e^a\wedge e^b.
\end{equation}
The construction of the five-dimensional Einstein--Hilbert action by means of the spin connection~\eqref{DRiemannconnection}, leads to the original KK theory.

Nevertheless, in a Riemann--Cartan geometry $M_N$ is not entirely described by $\hat{\tilde{\omega}}^{AB}$, but by a more general spin connection, independent of the metric degrees of freedom (see Appendix~\ref{Riemann-Cartan} for details). We can assume a general connection 1-form of the same type (regarding its $M_{N-1}\times S^1$ decomposition) as in Eq.~\eqref{DRiemannconnection}. Thus, the most general spin connection on $M_N$ compatible with the KK decomposition is given by 
\begin{equation}\label{Nconnection}
  \hat{\omega}^{AB} \equiv
  \begin{pmatrix}
    \omega^{ab}+\alpha^{ab}\hat{e}^N & \beta^a+\gamma^a\hat{e}^N\\
    -\beta^b-\gamma^b\hat{e}^N & 0
  \end{pmatrix}.
\end{equation}

@@

Similarly from the definition of torsion in Eq.~\eqref{tordef}, we find its distinctive parts to be
\begin{align}\label{T1}
  \hat{T}^a &= T^a+\left(\beta^a-\alpha^a_{\ \ b}e^b\right)\wedge\hat{e}^N,\\
  \label{T2}
  \hat{T}^N &= \sqrt{\phi}F-\beta_b\wedge e^b+\left(\frac{1}{2}\mbox{d}\ln\phi+\gamma_be^b\right)\wedge\hat{e}^N.
\end{align}

@@

Considering the Bianchi identities for the KK structure described above, we find relevant information about the new fields. Taking the exterior covariant derivative over the $N$-dimensional curvature and torsion, the Bianchi identities are
\begin{equation}
  \hat{\text{D}}\hat{R}^{AB} = 0 \quad \text{and} \quad \hat{\text{D}}\hat{T}^A = \hat{R}^A_{\ \ B}\wedge\hat{e}^B.
\end{equation}
A careful decomposition of the first one into its distinctive parts, gives the Bianchi identity for the curvature of the reduced spacetime, together with the second derivative rules
\begin{equation}
  \begin{aligned}
    \text{D}R^{ab} &=0, & \text{D}\mbox{D}\alpha^{ab} &=R^a_{\ \ l}\alpha^{lb}+R^b_{\ \ l}\alpha^{al},\\
    \text{D}\text{D}\beta^a &= R^a_{\ \ b}\beta^{b}, & \text{D}\text{D}\gamma^a &= R^a_{\ \ b}\gamma^{b}.
  \end{aligned}
\end{equation}
The second Bianchi identity, gives nothing but its equivalent for the manifold $M_{N-1}$, this is
\begin{equation*}
  \mbox{D}T^a=R^a_{\ \ b}\wedge e^b.
\end{equation*}

This is taken as a proof of the tensorial nature of these new fields under the four-dimensional Lorentz transformations.

@@

The most general theory requiring the Lagrangian to be (i)~an invariant $N$-form under local Lorentz transformations; (ii)~a local polynomial of the vielbein, the Lorentz connection and their exterior derivatives; (iii)~constructed without the Hodge dual,\footnote{The Hodge dual maps $p$-forms into $(N-p)$-forms through $\star\left(\hat{e}^{A_1}\wedge ... \wedge\hat{e}^{A_p}\right) = \frac{1}{(N-p)!}\hat{\epsilon}^{A_1\ldots A_p}{}_{A_{p+1}...A_N}\,\hat{e}^{A_{p+1}}\wedge ... \wedge\hat{e}^{A_N}$.} is the Lovelock--Cartan theory of gravity~\cite{Mardones:1990qc}. This is the natural generalization of Lanczos--Lovelock when torsional degrees of freedom are present.

>>>>

\author{Francisca \surname{Ram\'irez}.}
\affiliation{\UTFSM.}

\begin{abstract}
  We study the Kaluza--Klein dimensional reduction of the Lovelock--Cartan theory on a five-dimensional manifold, where the compact dimension has $S^1$ topology. We find cosmological solutions of the Friedmann--Robertson--Walker class for the effective theory. The torsional degrees of freedom induce a nonvanishing energy-momentum tensor in four dimensions. We find solutions describing expanding, contracting and bouncing universes. The model shows a dynamical compactification of the extra dimension in some regions of the parameter space. 
\end{abstract}

@@

In four dimensions, the Einstein--Hilbert action with cosmological constant is the most general theory which leads to second order field equations for the metric. In higher dimensions, however, particular combinations of higher order terms in the curvature can be added to the gravitational action, giving rise to second order field equations for the metric. The most general theory in arbitrary dimensions, which preserves the features of the four-dimensional Einstein--Hilbert, is described by the Lanczos--Lovelock action~\cite{Lanczos:1938sf,*Lovelock:1971yv}. Such a theory, has no ghosts~\cite{Zumino:1985dp} and it has the same degrees of freedom as the Einstein--Hilbert Lagrangian in arbitrary dimensions~\cite{Henneaux:1990au}.

@@

The Gauss--Bonnet term, in four dimensions, adds no dynamics to the metric, since it represents a topological invariant proportional to the Euler characteristic class, which can be written locally as a boundary term. Nevertheless, in dimensions higher than four it contributes to the field equations. This Lagrangian was also identified as the low-energy correction for a spin two field in string theory~\cite{Zwiebach:1985uq}.

The Riemannian five-dimensional Lanczos--Lovelock theory has been widely studied in the literature. For instance, exact wormhole solutions have been found in vacuum~\cite{Dotti:2006cp,*Dotti:2007az} and coupled with matter fields~\cite{Mehdizadeh:2015jra}. In higher order Lanczos--Lovelock models, this class of exact solutions have been reported in~\cite{Mehdizadeh:2015dta} and in higher order compactified Lanczos--Lovelock theories with torsion in~\cite{Canfora:2008ka}. Additionally, the compactification of higher Lanczos--Lovelock terms has been considered in Refs.~\cite{MuellerHoissen:1985mm,*MuellerHoissen:1989yv} and their cosmology in Refs.~\cite{MuellerHoissen:1985ij,Deruelle:1986iv,Deruelle:1989fj}. It is worth mentioning that, classically, wormholes must be supported by a kind energy compatible with the cosmological hypothesis. Their existence in these extended models implies the presence of degrees of freedom that could provide an explanation for the exotic matter/energy abundance in the Universe.

On the other hand, it is well-known that GR assume the torsion-free condition {\it a priori}. However, as E.~Cartan first considered, it is possible to take metricity and parallelism as truly independent concepts~\cite{Cartan1922,*Cartan1924,*Cartan1925}. Such kind of geometries are known as Riemann--Cartan and offer the natural framework for Poincar\'e gauge theories, where the torsion appears as the field strength of translations, sourced by the spin-current~\cite{Kibble:1961ba,Sciama:1962,Hehl:1976kj}.
%In such a framework, the vielbein is interpreted as the gauge connection of translations and the torsion as its field strength. The formulation of gravity as a gauge theory has major benefits in the problem of a unified picture of interactions, since it allows to interpret it as arising through a localization procedure, where the spacetime is a Riemann--Cartan manifold. 
%Due to this, several authors have relaxed the torsion-free condition since, as E. Cartan argued, there is no a fundamental reason to assume a torsion-free condition {\it a priori}. 
Moreover, the vacuum predictions of its simplest formulation---the Einstein--Cartan theory---holds the experimental tests of GR. The cosmological consequences of non-Riemannian geometries has been widely studied in the literature (for a review, see Ref.~\cite{Puetzfeld:2004yg}). %For instance, the pioneer works of Kopczy\'nski showed how the introduction of a spinning dust, acting as a source of the curvature and torsion, avoids the appearance of cosmological singularities~\cite{Kopczynski:1972,Kopczynski:1973}. In the context of the early evolution of the Universe, the spin density can drive the inflationary phase within the Einstein--Cartan theory, along with the avoidance of singularities~\cite{Gasperini:1986mv}. It has been shown in Ref.~\cite{Shie:2008ms} that torsional degrees of freedom might produce the current phase of accelerated expansion of the Universe. Additionally, the nonminimal coupling of the Euler characteristic form with a scalar field in Riemann--Cartan spacetimes, allows the torsional degrees of freedom to behave as an exotic induced-matter content, customary for the phase of accelerated expansion~\cite{Toloza:2013wi}.

In the framework of the Kaluza--Klein theories in higher dimensional Riemann--Cartan geometries, phenomenology of the extra dimensional torsion was found in Ref.~\cite{Kalinowski:1980da}, metric dependent torsion in extra dimensions was studied in~\cite{Shankar:2012vd}, along with its consequences in cosmology~\cite{Chen:2009ep}. Compatification of higher dimensional Brans--Dicke models with torsion was considered in Ref.~\cite{German:1993bq}. Torsion-free black-hole solutions were found in~\cite{Aros:2007nn}, for first order compactified gravity.

The extension of the Lanczos--Lovelock theories with nonvanishing torsion is known as Lovelock--Cartan~\cite{Mardones:1990qc}. In that framework, we present a new class of cosmological solutions in five dimensions, where the compact dimension is $S^1$. The theory admits a nonvanishing torsion in vacuum due to the presence of the Gauss--Bonnet term, in contrast to the five-dimensional Einstein--Cartan theory. After the dimensional reduction, the extra degrees of freedom generate an energy density and pressure as four-dimensional matter fields. In some cases, the solutions avoid the appearance of initial singularities and drive the accelerated expansion of the Universe, while the radius of the compact manifold goes to zero.

This work is organized as follows: In Sec.~\ref{KK} we review the Kaluza--Klein geometry in the first order formalism and fix our notation and conventions. In Sec.~\ref{5EGB}, we study the dynamics of the general Lovelock--Cartan action in five-dimensional spacetime and its dimensional reduction. In Sec.~\ref{cosmos}, we look for cosmological solutions of the Friedmann--Robertson--Walker class with nontrivial torsion. Conclusions and remarks are given in Sec.~\ref{conclusions}. We also incorporate a number of appendices to make this work a self-contained article.

\section{Kaluza--Klein Geomatry in Riemann--Cartan Spacetimes\label{KK}}

In the following we will consider $M_N$ to be a $N$-dimensional differential manifold. Every quantity defined on $M_N$ will be denoted by hats $\hat{x}$. Capital Greek characters (coordinate indices) and capital Latin characters (Lorentz indices) run over the $N$ dimensions, i.~e. $A = 0,...,N-2,N$, while lower case run in the $(N-1)$-dimensional reduced manifold, i.~e. $a = 0,...,N-2$.

The differential structure of $M_N$ is determined by two fields. The spin connection 1-form, $\hat{\omega}^{AB} = \hat{\omega}^{AB}{}_{\Gamma}\,\text{d}\hat{x}^\Gamma$, describes the affine structure of the $N$-dimensional manifold and the vielbein 1-form, $\hat{e}^A=\hat{e}^{A}{}_{\Gamma}\,\text{d}\hat{x}^\Gamma$, defines the metric structure of the same manifold through the relation $\hat{g}_{\Gamma\Delta} = \hat{e}^{A}{}_{\Gamma}\hat{e}^{B}{}_{\Delta}\hat{\eta}_{AB}$, where $\hat{g}_{\Gamma\Delta}$ is the spacetime metric, while $\hat{\eta}_{AB} = \diag(-,+,...,+)$ is the Minkowski metric. The vielbein maps coordinate into Lorentz indices.
In our convention, the Levi-Civita symbol is such that $\hat{\epsilon}_{01...(N-2)N} = 1$ and the reduced symbol is defined by fixing the last index, i.~e.
\[\hat{\epsilon}_{a_1 \cdots a_{N-1} N} \equiv \epsilon_{a_1 \cdots a_{N-1}}.\]
%% \[\hat{\epsilon}_{0 1 \cdots N-1} = {\epsilon}_{0 1 \cdots N-1} = 1,\]
Additionally, all the Riemannian fields (torsion-free) will be explicited by a tilde, as it was done in Eq.~\eqref{GB} for the curvature.
%% In our convention, the flat Lorentzian metric will be $\hat{\eta}_{AB}=\diag(-,+,\cdots,+)$, and the $N$-dimensional Levi-Civita symbol $\hat{\epsilon}_{A_1 \cdots A_N}$ is reduced by defining
%% \begin{align*}
%% \hat{\epsilon}_{a_1 \cdots a_{N-1} N} \equiv \epsilon_{a_1 \cdots a_{N-1}},
%% \end{align*} 
%% since the coordinates of $M_N$ run from $0$ to $N-1$. Our convention fixes $\hat{\epsilon}_{01...N}=1$. Hereon, Riemannian fields (torsion-free) will be explicited by a tilde, as it was done in Eqs.~\eqref{GB} for the curvature.

Due to the topology of the manifold, we can expand the depenence of the fields on the extra coordinate, $z$, in Fourier series as  
\begin{equation}
  \label{Fourier}
  \hat{\varrho}(\hat{x}^\Gamma)=\sum_n\varrho_{(n)}(x)e^{i n z},
\end{equation}
where $x$ denotes the coordinates of the $(N-1)$-manifold, collectively. Henceforth, we will focus on the $n=0$ mode of the expasion, also refered to as the low-energy sector.

@@

The Kaluza--Klein (KK) ansatz for the metric lies on the premise that the compact dimension of $M_N$ is orthogonal to the rest of the manifold at each point. This leads to the following metric structure
\begin{equation}
  \hat{g}_{\Gamma\Delta} =
  \begin{pmatrix}
    g_{\gamma\delta} +\frac{\hat{g}_{\gamma z}\hat{g}_{\delta z}}{\hat{g}_{zz}}&\hat{g}_{\gamma z}\\
    \hat{g}_{z\delta} & \hat{g}_{zz}
  \end{pmatrix}
  =
  \begin{pmatrix}
    g_{\gamma\delta} + \phi A_\gamma A_\delta&\phi A_\gamma\\
    \phi A_{\delta} & \phi
  \end{pmatrix},
\end{equation}
and its inverse
\begin{equation}
  \hat{g}^{\Gamma\Delta}=
  \begin{pmatrix}
    g^{\gamma\delta}&-A^\gamma\\
    -A^{\delta} & \phi^{-1}+A^2
  \end{pmatrix}.
\end{equation}
It introduces a scalar field $\phi$ and a vector field $A_\mu$ as new gravitational degrees of freedom. The vielbein which holds this structure for the metric has the following form
\begin{equation}
  \label{Dvielbein}
  \hat{e}^A{}_{\Gamma} =
  \begin{pmatrix}
    \hat{e}^a{}_{\gamma}& 0\\
    \hat{e}^N{}_{\gamma} & \hat{e}^N{}_{z}
  \end{pmatrix}
  =
  \begin{pmatrix}
    e^a{}_{\gamma}& 0\\
    \sqrt{\phi}A_\gamma & \sqrt{\phi}
  \end{pmatrix},
\end{equation}
and its inverse, $\hat{E}_A{}^{\Gamma}$, defined such that $\hat{E}_A{}^{\Gamma}\hat{e}^A{}_{\Delta}=\delta^\Gamma_{\Delta}$ and $\hat{E}_A{}^{\Gamma}\hat{e}^B{}_{\Gamma}=\delta^A_B$, reads
\begin{equation}
  \label{Dinversevielbein}
  \hat{E}_A{}^{\Gamma} =
  \begin{pmatrix}
    \hat{E}_a{}^{\gamma}& 0\\
    \hat{E}_a{}^{z} & \hat{E}_N{}^{z}
  \end{pmatrix}
  =
  \begin{pmatrix}
    E_a{}^{\gamma}& 0\\
    -A_a & \sqrt{\phi}^{-1}
  \end{pmatrix},
\end{equation}
where $A_a = E_a{}^\mu A_\mu$. This shows that the compact manifold $S^1$ has its own tangent space, $T_pS^1$, at each point $p\in M_N$, and is independent of $T_pM_{N-1}$ in the sense that they do not mix. %It seems reasonable to think that if we want two manifolds to be orthogonal to one another, they will not mix vectors after the tangent mapping.

@@

A Riemannian connection compatible with the \mbox{$N$-dimensional} vielbein~\eqref{Dvielbein}, is built under the premise that $\mbox{d}\hat{e}^A + \hat{\tilde{\omega}}^{A}{}_B \wedge \hat{e}^B = 0$. Thus, we find
\begin{equation}
  \label{DRiemannconnection}
  \begin{split}
    \hat{\tilde{\omega}}^{ab}&=\tilde{\omega}^{ab}-\frac{1}{2}\sqrt{\phi}F^{ab}\hat{e}^N,\\
    \hat{\tilde{\omega}}^{Na}&=\frac{1}{2}\sqrt{\phi}F^a_{\ \ l}e^l+\frac{1}{2}\partial^a\ln\phi\hat{e}^N,
  \end{split}
\end{equation}
where $\tilde{\omega}^{ab}$ is the Riemannian spin connection of the reduced manifold and $F_{ab}$ are the components of the field strength of $A = A_m\, e^m$, defined by
\begin{equation}
  F=\text{d}A=\frac{1}{2}F_{ab}\, e^a\wedge e^b.
\end{equation}
The construction of the five-dimensional Einstein--Hilbert action by means of the spin connection~\eqref{DRiemannconnection}, leads to the original KK theory.

In a Riemann--Cartan geometry $M_N$ is not entirely described by $\hat{\tilde{\omega}}^{AB}$, but by a more general spin connection independent of the metric degrees of freedom (see Appendix~\ref{Riemann-Cartan} for details). We can assume a general connection 1-form of the same type (regarding its $M_{N-1}\times S^1$ decomposition) as in Eq.~\eqref{DRiemannconnection}. Thus, the most general spin connection on $M_N$ compatible with the KK decomposition is given by 
\begin{equation}\label{Nconnection}
  \hat{\omega}^{AB} \equiv
  \begin{pmatrix}
    \omega^{ab}+\alpha^{ab}\hat{e}^N & \beta^a+\gamma^a\hat{e}^N\\
    -\beta^b-\gamma^b\hat{e}^N & 0
  \end{pmatrix}.
\end{equation}

@@

Similarly from the definition of torsion in Eq.~\eqref{tordef}, we find its distinctive parts to be~\footnote{Henceforth we will refer to as distinctive parts regarding the $M_{N-1}\times S^1$ decomposition.}
\begin{align}\label{T1}
  \hat{T}^a &= T^a+\left(\beta^a-\alpha^a_{\ \ b}e^b\right)\wedge\hat{e}^N,\\
  \label{T2}
  \hat{T}^N &= \sqrt{\phi}F-\beta_b\wedge e^b+\left(\frac{1}{2}\mbox{d}\ln\phi+\gamma_be^b\right)\wedge\hat{e}^N.
\end{align}

@@

Considering the Bianchi identities for the KK structure described above, we find relevant information about the new fields. Taking the exterior covariant derivative over the $N$-dimensional curvature and torsion, the Bianchi identities are
\begin{equation}
  \hat{\text{D}}\hat{R}^{AB} = 0 \quad \text{and} \quad \hat{\text{D}}\hat{T}^A = \hat{R}^A_{\ \ B}\wedge\hat{e}^B.
\end{equation}
A careful decomposition of the first one into its distinctive parts, gives the Bianchi identity for the curvature of the reduced spacetime, together with the second derivative rules
\begin{equation}
  \begin{aligned}
    \text{D}R^{ab} &=0, & \text{D}\mbox{D}\alpha^{ab} &=R^a_{\ \ l}\alpha^{lb}+R^b_{\ \ l}\alpha^{al},\\
    \text{D}\text{D}\beta^a &= R^a_{\ \ b}\beta^{b}, & \text{D}\text{D}\gamma^a &= R^a_{\ \ b}\gamma^{b}.
  \end{aligned}
\end{equation}
This is taken as a proof of the tensorial nature of these new fields under the four-dimensional Lorentz transformations. The second Bianchi identity gives nothing but its equivalent for $M_{N-1}$, this is
\begin{equation*}
  \mbox{D}T^a=R^a_{\ \ b}\wedge e^b.
\end{equation*}

@@

The most general theory requiring the Lagrangian to be (i)~an invariant $N$-form under local Lorentz transformations; (ii)~a local polynomial of the vielbein, the Lorentz connection and their exterior derivatives; (iii)~constructed without the Hodge dual,\footnote{The Hodge dual maps $p$-forms into $(N-p)$-forms through $\star\left(\hat{e}^{A_1}\wedge ... \wedge\hat{e}^{A_p}\right) = \frac{1}{(N-p)!}\hat{\epsilon}^{A_1\ldots A_p}{}_{A_{p+1}...A_N}\,\hat{e}^{A_{p+1}}\wedge ... \wedge\hat{e}^{A_N}$.} is the Lovelock--Cartan theory of gravity~\cite{Mardones:1990qc}. This is the natural generalization of Lanczos--Lovelock action when torsional degrees of freedom are present. Its simplest realization is the Einstein--Cartan model and their dimensional reduction is indistinguishable from the Riemannian case (see for instance~\cite{German:1993bq,Aros:2007nn}).

In five dimensions, the Lovelock--Cartan theory is given by the following action principle 
\begin{widetext}
  \begin{equation}
    \label{action5EGB}
    I = \int_{M_5} \hat{\epsilon}_{ABCDE} \Big(\frac{\alpha_0}{5}\hat{e}^A\wedge\hat{e}^B\wedge\hat{e}^C\wedge
    \hat{e}^D\wedge\hat{e}^E
    +\frac{\alpha_1}{3}\hat{R}^{AB}\wedge\hat{e}^C\wedge\hat{e}^D\wedge\hat{e}^E
    +\alpha_2\hat{R}^{AB}\wedge\hat{R}^{CD}
    \wedge\hat{e}^E\Big),
  \end{equation}
  where $\alpha_0$, $\alpha_1$ and $\alpha_2$ are dimensionful coupling constants. Its variation with respect to the five-dimensional vielbein and spin connection give the equations
  \begin{gather}
    \label{delta e}
    \hat{\epsilon}_{ABCDE}\Big(\alpha_0\hat{e}^A\wedge\hat{e}^B\wedge\hat{e}^C\wedge\hat{e}^D
    + \alpha_1\hat{R}^{AB}\wedge\hat{e}^C\wedge\hat{e}^D
    + \alpha_2\hat{R}^{AB}\wedge\hat{R}^{CD}\Big)=0,
    \\
    \label{delta w}
    \hat{\epsilon}_{ABCDE}\Big(\alpha_1\hat{e}^C\wedge\hat{e}^D+
    2\alpha_2\hat{R}^{CD}\Big)\wedge\hat{T}^E=0.
  \end{gather}
\end{widetext}
